Bird beak's elaboration and exploration: data preparation

### 2021-10-12 Analysis/01-Data_preparation.Rmd
fixed levels names (to match the ones in the dataset!)

```{r install_beer, eval = FALSE}
## Installing beer
devtools::install_github("TGuillerme/elaboration_exploration_bird_beaks/beer")
```

```{r load_beer, eval = TRUE}
## Loading packages
library(dispRity)
library(beer)
library(knitr)
```

@@

```{r}
## Reading the clades list
clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
## Remove the absent species
clades_list <- clades_list[rownames(clades_list) %in% rownames(shapespace), ]

#@param table the clade list data frame
#@param level the ID of the column to use as a level
#@param threshold drop levels with less than the threshold
make.level.list <- function(table, level, threshold = 20) {
  ## Select the level
  table_level <- table[, level]
  ## Find the empty level
  empty_level <- which(levels(table_level) == "")
  if(length(empty_level) != 0) {
      ## Remove the empties
      level_list <- as.factor(as.character(table_level[-which(table_level == "")]))
  }
  empty_level <- ""
  ## Select the levels below the threshold
  remove_levels <- which(table(level_list) < threshold)
  ## Remove the levels
  while(length(remove_levels) > 0) {
    level_list <- as.factor(as.character(level_list[-which(level_list == names(remove_levels)[1])]))
    table_level[which(table_level == names(remove_levels)[1])] <- empty_level
    table_level <- as.factor(as.character(table_level))
    remove_levels <- remove_levels[-1]
  }
  ## Convert into a list
  factors <- levels(level_list)
  list_levels <- list()
  while(length(factors) > 0) {
    list_levels[[length(list_levels) + 1]] <- which(level_list == factors[1])
    factors <- factors[-1]
  }
  ## Name the elements correctly
  names(list_levels) <- levels(level_list)

  ## Also returns the new column
  return(list(col = table_level, list = list_levels))
}

## Reading the clades list
groups_level1 <- make.level.list(clades_list, level = 1, threshold = 20)
groups_level2 <- make.level.list(clades_list, level = 2, threshold = 20)
all_groups <- c(groups_level1$list, groups_level2$list)
group_names_level_1_2 <- all_groups
save(group_names_level_1_2, file = "../Data/Processed/group_names_level_1_2.rda")
```

@@

```{r}
## Make a dispRity object and get axes data
shape_axes <- select.axes(custom.subsets(shapespace, group = all_groups))
form_axes  <- select.axes(custom.subsets(formspace, group = all_groups))
```

```{r, echo = FALSE}
kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X1")
```

```{r, echo = FALSE}
kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X2")
```

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(shape_axes)
```
Fig X1: variance per axes for the Shape space

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(form_axes)
```
Fig X1: variance per axes for the Form space

>>>>

```{r install_beer, eval = FALSE}
## Installing the dispRity branch
devtools::install_github("TGuillerme/dispRity@MCMCglmm")
```

```{r load_beer, eval = TRUE}
## Loading packages
library(dispRity) #v>=1.7
```

@@

```{r}
## Reading the clades list
clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
## Remove the absent species
clades_list <- clades_list[rownames(clades_list) %in% rownames(shapespace), ]
## Sorting the list by rownames from the shapespace
clades_list <- clades_list[match(rownames(shapespace), rownames(clades_list)), -c(6, 7)]
## Renaming the levels
colnames(clades_list) <- c("superorder", "order", "suborder", "superfamily", "family")

## Cleaning the levels (removing any level with less than 15 elements)
update.level <- function(level, min = 15) {
    to_remove <- which(table(level) < min)
    remove_table <- data.frame("clade" = names(to_remove), "species" =  c(table(level)[to_remove]))
    level[which(level %in% c(names(to_remove)))] <-  ""
    return(list(level = as.factor(as.character(level)), removed = remove_table))
}

## Updated list
update_clade_list <- lapply(as.list(clades_list), update.level, min = 15)
clade_list <- data.frame(lapply(update_clade_list, `[[`, 1))
removed_table <- do.call(rbind, lapply(update_clade_list, function(x) {rownames(x$removed) <- NULL; return(x$removed)}))

## Saving the clade list
save(clade_list, file = "../Data/Processed/clade_list.rda")
```

```{r, echo = FALSE}
library(knitr)
kable(removed_table, caption = "List of removed clades by level and number of species in that clade in the dataset.", label = "Table X1")
```

@@

```{r}
## Temporary change the empty levels to noclade
clades <- clades_list[, c(1)]
levels(clades)[1] <- "noclade"

## Make a dispRity object and get axes data
shape_axes <- select.axes(custom.subsets(shapespace, group = clades))
form_axes  <- select.axes(custom.subsets(formspace, group = clades))
```

```{r, echo = FALSE}
kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X2")
```

```{r, echo = FALSE}
kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X3")
```

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(shape_axes)
```
Fig X1: variance per axes per superorder and order in the shapespace

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(form_axes)
```
Fig X1: variance per axes per superorder and order in the formspace
